Periodograms for Multiband Astronomical Time Series

### 2015-01-24 writeup/paper.tex
more paper work; add appendix

@@

\todo{In \S 2, we provide a brief overview of periodograms, including the matrix-based formulation used in this work. In \S...}

@@

Below, we'll apply the simplest of this class of models, the $(1, 0)$ shared-phase model, to data from the SDSS-2 survey.

@@

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig08a.pdf}
  \includegraphics[width=\textwidth]{fig08b.pdf}
  \caption{
    This figure shows a repeat of the results from \fig{compare_periods}, but the data is artificially reduced to only a single-band observation on each evening, a situation reflective of the observing cadence of future large-scale surveys (compare to \fig{compare_periods}).
    In this case, the single-band SuperSmoother strategy used by \citet{Sesar2010} fails: there is simply not enough data in each band to recover an accurate period estimate, and the correct period is among the top 5 candidates in fewer than 60\% of cases.
    The shared-phase multiband approach utilizes information from all five bands, and returns much more robust results: even with the reduced data, the true period is among the top 5 candidates in 94\% of cases.
  } 
  \figlabel{compare_periods_partial}
\end{figure}

\todo{Discuss the results in \fig{compare_periods} and \fig{compare_periods_partial}}

@@

\todo{Discuss issues with aliasing \& window function corrections}

\todo{Discuss issues with physicality of the model?}

@@

%\appendix

>>>>

% For Python code
\usepackage{listings}
\definecolor{lbcolor}{rgb}{0.9,0.9,0.9}
\lstset{language=Python,
        basicstyle=\footnotesize\ttfamily,
        showspaces=false,
        showstringspaces=false,
        tabsize=2,
        breaklines=false,
        breakatwhitespace=true,
        identifierstyle=\ttfamily,
        keywordstyle=\bfseries\color[rgb]{0.133,0.545,0.133},
        commentstyle=\color[rgb]{0.133,0.545,0.133},
        stringstyle=\color[rgb]{0.627,0.126,0.941},
    }

@@

\todo{In \S 2, we provide a brief overview of periodograms, including the matrix-based formulation used in this work. In \S... \citet[][hereafter S10]{Sesar2010}...}

@@

Below, we'll apply the simplest of this class of models, the $(1, 0)$ shared-phase model, to data from the Stripe 82 of the Sloan Digital Sky Survey.

@@

Stripe 82 is a three hundred square degree equatorial region of the sky which was repeatedly imaged through multiple band-passes during the second phase of the Sloan Digital Sky Survey (SDSS II).
As a real-world test of the multiband periodogram method, we consider the SDSS-2 observations of 483 RR Lyrae stars compiled and studied by S10.
S10 determined periods for these stars based on an emperically-derived light curve templates. Because the template-fitting method is extremely computationally intensive, S10 first determined candidate periods by taking the top 5 results of the supersmoother \citep{Reimann94} algorithm applied to the $g$-band; template fits were then performed at each candidate period and the period with the best template fit was reported as the true period. In this section, we make use of this dataset to quantitatively evaluate the effectiveness of the multiband periodogram approach in the case of this real-world data.

\subsection{Densely-sampled Multiband Data}

@@

The full S10 RR Lyrae dataset consists of 483 objects with an average of 55 observations in each of the five SDSS $ugriz$ bands spread over just under ten years.  In the upper panels of \fig{compare_periods} we show the observed data for one of these objects, along with the periodogram derived with the single-band supersmoother algorithm and the shared-phase $(0, 1)$-multiband algorithm\footnote{The supersmoother periodogram $P_{SS}$ is constructed from the minimum sum of weighted model residuals $\bar{r}_{min}$ in analogy with the Lomb-Scargle periodogram's $\chi^2$ interpretation from \eq{chi2PN}: $P_{SS}(\omega) = 1 - \bar{r}_{min}(\omega) / \bar{r}_0$}. Here we have a case which is analogous to that shown for simulated data in the top panels of \fig{multiband_sim}: each band has enough data to easily locate candidate peaks, the best of which is selected via the S10 template-fitting procedure.

The lower panels of \fig{compare_periods} compare the S10 period with the best periods obtained from the 1-band supersmoother (lower-left) and from the shared-phase multiband model (lower-right). To guide the eye, the figure includes indicators of the locations of beat aliases (dotted lines) and multiplicative aliases (dashed lines) of the S10 period.

The best-fit supersmoother period matches the S10 period in 89\% of cases (421/483), while the best-fit multiband period matches the S10 period in 79\% of cases (382/483). The modes of failure are instructive: when the supersmoother model misses the S10 period, it tends to land on a multiplicative alias. This is due to the flexibility of the supersmoother model: a doubled period spreads the points out further, leading to fewer constraints in each neighborhood and thus a smaller average residual around the over-fit model.  When the multiband model misses the S10 period, it tends to land on a beat alias between the S10 period and the 1-day observing cadence. This is due to the fact that the single-frequency periodic model significantly under-fits the data, and thus cannot distinguish deviations due to underfitting from deviations due to window function effects.

In both models, the S10 period appears among the top 5 periods 99\% of the time: $477/483$ for supersmoother, and $480/483$ for multiband.\footnote{We would expect this correspondence to be 100\% in the case of the $g$-band supersmoother, which was the model used in the first pass of the S10 computation. This discrepancy can be chalked-up to the different supersmoother implementations used in S10 and in this work. The discrepant curves are those with very low signal-to-noise.} This suggests that had S10 used the multiband Lomb-Scargle rather than the supersmoother in the first pass for that study, the final results presented there would be for the most part unchanged.

The results of this subsection show that the shared-phase multiband approach is comparable to the supersmoother approach for densely-sampled multiband data, although it has a tendency to get fooled by structure in the survey window function.

\subsection{Sparsely-sampled Multiband Data}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{fig08a.pdf}
  \includegraphics[width=\textwidth]{fig08b.pdf}
  \caption{
    This figure shows a repeat of the results from \fig{compare_periods}, but the data is artificially reduced to only a single-band observation on each evening, a situation reflective of the observing cadence of future large-scale surveys (compare to \fig{compare_periods}).
    In this case, the single-band SuperSmoother strategy used by \citet{Sesar2010} fails: there is simply not enough data in each band to recover an accurate period estimate, and the correct period is among the top 5 candidates in fewer than 60\% of cases.
    The shared-phase multiband approach utilizes information from all five bands, and returns much more robust results: even with the reduced data, the true period is among the top 5 candidates in 94\% of cases.
  } 
  \figlabel{compare_periods_reduced}
\end{figure}

Where we expect the multiband approach to have an advantage is when the data are sparsely sampled, with observations through only a single band each night. To simulate this, we reduce the Stripe 82 RR Lyrae dataset by a factor of 5, keeping only a single band of imaging each night, for an average of 11 observations of each object per band. This is much closer to the type of observation available in future multiband surveys.

The upper panels of \fig{compare_periods_reduced} show an example light curve from this reduced dataset, along with the supersmoother and multiband periodograms derived from this data. Analogously to the lower panels of \fig{multiband_sim}, the single-band supersmoother model loses the true period within the noise, while the shared-phase multiband model still shows a noticeable spike at this true period.

The lower panels of \fig{compare_periods_reduced} show the relationship between the S10 periods (based on the full dataset) and the periods derived with each model from this reduced dataset. It is clear that the supersmoother model is simply over-fitting noise with this few data points: the top period matches S10 in only 32\% of cases (compared to 87\% with the full dataset), and the top 5 periods contain the S10 period only 57\% of the time. The failure mode is much less regular as well: rather than being clustered near aliases, most of the points are scattered seemingly randomly around the space.

The multiband method does much better on the reduced dataset. Even with an 80\% reduction in the number of observations, the multiband method matches the S10 period 66\% of the time (compared to 79\% with the full dataset), and the top 5 periods contain the S10 period 94\% of the time. This performance is due to the fact that the multiband algorithm has relatively few parameters, but is yet able to accommodate data from multiple observing bands. In particular, this suggests that with the multiterm periodogram, the S10 analysis could have been done effectively with only a small fraction of the available data. This bodes well for future surveys, where data on variable stars will be much more sparsely sampled.

@@

\todo{Zeljko, can you take a stab at this?}

\begin{enumerate}
  \item Our method kicks ass, therefore we kick ass. QED.
  \item Future work: account for beat-aliases for ground-based data
  \item Future work: physically-motivated priors to further simplify the model
  \item Opportunity to apply this to PanSTARRS data release
\end{enumerate}

@@

\appendix
\section{{\tt multiband\_LS}: Python Implementation of Multiband Periodogram}
The algorithm outlined in this paper is available as an open-source Python package\footnote{\url{http://github.com/jakevdp/multiband_LS/}}, along with code to download the data and recreate all the figures in this paper.
{\tt multiband\_LS} is a pure-Python package written to be compatible with both Python 2 and Python 3, and performs fast numerical computation via the {\tt numpy} package\footnote{http://www.numpy.org}.

The API for the module is largely influenced by that of the {\tt scikit-learn} package\footnote{\url{http://scikit-learn.org}} \citep{scikit-learn}, in which models are class objects which can be ``fit()`` to data.
Here is a basic example of how you can use the code to download data, fit a multiband model to the data, and compute the power at a few periods:

\begin{lstlisting}
from multiband_LS import LombScargleMultiband

# Fetch the Sesar 2010 RR Lyrae data
from multiband_LS.data import fetch_light_curves
data = fetch_light_curves()
t, mag, dmag, filts = data.get_lightcurve(data.ids[0],
                                          return_1d=True)

# Construct the multiband model
model = LombScargleMultiband(Nterms_base=0, Nterms_band=1)
model.fit(t, mag, dmag, filts)

# Compute power at the following periods
model.periodogram([0.2, 0.3, 0.4])
\end{lstlisting}

Other models are available as well. For example, here is how you can compute the periodogram under the supersmoother model:

\begin{lstlisting}
from multiband_LS import SuperSmoother

# Construct the supersmoother model
model = SuperSmoother()
gband = (filts == 'g')
model.fit(t[gband], mag[gband], dmag[gband])

# Compute power at the following periods
model.periodogram([0.2, 0.3, 0.4])
\end{lstlisting}

The code contains many more methods, and much more functionality that what is shown here. For updates and more information, visit \url{http://github.com/jakevdp/multiband_LS}.
